\documentclass[10pt]{article}

\usepackage[margin=0.85in]{geometry}
\usepackage{times}
\usepackage{hyperref}
\usepackage{url}
\usepackage{enumitem}
\usepackage{titlesec}

% Compact spacing
\setlength{\parskip}{4pt}
\setlength{\parindent}{0pt}
\titlespacing*{\section}{0pt}{10pt}{4pt}
\titlespacing*{\subsection}{0pt}{8pt}{3pt}

\title{\vspace{-0.6in}Do Robots Need Cognitive Memory? An Empirical Study of\\Multi-Tier LLM-Augmented Memory for Robotic Agents\vspace{-0.15in}}
\author{Alex Schott \and Leo Li}
\date{\vspace{-0.25in}}

\begin{document}
\maketitle

\section{Problem and Motivation}

LLM-controlled robotic systems where a language model directly selects actions from sensor input require memory to operate over extended time horizons. Without memory, a robot cannot recall past observations, learn from failed actions, or build a persistent model of its environment. Recent work on LLM agentic memory (MemGPT, Reflexion, Generative Agents) have made significant progress, but these systems are designed for text-native environments where observations arrive as natural language and at human-interactive rates. Robotic systems face fundamentally different constraints:

\textbf{Data volume and modality.} Robots produce continuous, multi-modal sensor data (RGB, LiDAR, joint states) at 10--100\,Hz---far exceeding what any context window can hold, and in formats that require interpretation before an LLM can reason about them.

\textbf{Latency sensitivity.} When the LLM is on the critical path of the control loop, memory retrieval latency directly degrades robot responsiveness. Text-based agents can tolerate seconds of retrieval time; robots operating at ${\sim}1$\,Hz action frequency cannot.

\textbf{Compression requirement.} Raw sensor data must be translated into language, then progressively compressed and abstracted to remain useful over long time horizons.

These constraints suggest that naive approaches such as appending all observations to the context window, or flat vector retrieval over unprocessed data will fail at scale. A more structured approach may be necessary, but this has not been empirically validated.

\section{Related Work}

\textbf{LLM agent memory.} MemGPT \cite{packer2023memgpt} introduces hierarchical memory management for LLM agents with explicit memory tiers. Generative Agents \cite{park2023generative} uses a retrieval-based memory with reflection. Reflexion \cite{shinn2023reflexion} stores verbal feedback for self-improvement. These systems operate on text observations and do not address sensor-to-language translation or real-time latency constraints.

\textbf{Cognitive architectures for robots.} Recent work like SayCan \cite{ahn2022saycan} and TypeFly \cite{chen2023typefly} use LLMs for robot planning but do not incorporate persistent memory systems. STaR \cite{star2025} introduces periodic VLM captioning for robot observation but does not address memory consolidation or retrieval.

\textbf{Our positioning.} We bridge these lines of work by empirically studying whether cognitive-inspired memory decomposition (observation, consolidation, injection) provides measurable benefits for LLM-controlled robots over simpler alternatives.

\section{Approach}

We introduce \textbf{typemem}, a minimal abstract framework for robot memory that decomposes memory management into three function types operating over a generic vector store:

\textbf{Observation functions} run in the background, converting raw sensor data into natural language memories stored in the database.
%
\textbf{Consolidation functions} run in the background, reading existing memories and creating summaries, extracting patterns, or pruning redundant entries---often using LLM calls for reasoning-heavy operations.
%
\textbf{Injection functions} are called synchronously before each LLM inference, searching memory and selecting the most relevant items to include in the prompt within a token budget.

The memory store interface is deliberately minimal (add, search, delete, update) with no built-in opinion on tiers, retention, or linking. All organizational policy lives in the functions themselves, enabling controlled ablation.

A key feature is \textbf{LLM-assisted function generation}: given a natural language description of a robot's personality and use case, the system generates starter implementations (Python code for deterministic operations, LLM prompts for reasoning tasks) that users can review and refine.

\section{Experimental Plan}

We compare three strategies, all implemented as instances of the same typemem interface:

\begin{itemize}[nosep, leftmargin=1.5em]
    \item \textbf{Full context.} Observations appended as raw text, no consolidation, inject everything up to the token limit.
    \item \textbf{Monolithic RAG.} Observations embedded and stored, no consolidation, top-$k$ cosine retrieval at injection.
    \item \textbf{Tiered memory (ours).} LLM-generated observation, consolidation, and injection functions with structured memory organization.
\end{itemize}

\subsection{Evaluation}

\textbf{Latency benchmarks.} Measure injection time and consolidation overhead across memory store sizes (100 to 100K entries) for all three strategies.

\textbf{Synthetic benchmarks.} Scripted robot observation streams (e.g., ``patrol a building for 1 hour'') with ground-truth retrieval questions at decision points. Metrics: retrieval precision/recall, token efficiency, task-relevant information coverage.

\textbf{Real robot evaluation (stretch goal).} Deploy all three strategies on a Unitree Go2 quadruped running identical navigation and interaction tasks via the TypeGo platform. Measure task completion rate and decision quality.

\bibliographystyle{plain}
\begin{thebibliography}{10}

\bibitem{ahn2022saycan}
M.~Ahn, A.~Brohan, N.~Brown, Y.~Chebotar, O.~Cortes, B.~David, C.~Finn, C.~Fu, K.~Gopalakrishnan, K.~Hausman, et~al.
Do as {I} can, not as {I} say: Grounding language in robotic affordances.
In \emph{CoRL}, 2022.
\url{https://arxiv.org/abs/2204.01691}

\bibitem{packer2023memgpt}
C.~Packer, S.~Wooders, K.~Lin, V.~Fang, S.~G. Patil, I.~Stoica, and J.~E. Gonzalez.
{MemGPT}: Towards {LLMs} as operating systems.
\emph{arXiv preprint arXiv:2310.08560}, 2023.
\url{https://arxiv.org/abs/2310.08560}

\bibitem{park2023generative}
J.~S. Park, J.~C. O'Brien, C.~J. Cai, M.~R. Morris, P.~Liang, and M.~S. Bernstein.
Generative agents: Interactive simulacra of human behavior.
In \emph{UIST}, 2023.
\url{https://arxiv.org/abs/2304.03442}

\bibitem{shinn2023reflexion}
N.~Shinn, F.~Cassano, A.~Gopinath, K.~Narasimhan, and S.~Yao.
Reflexion: Language agents with verbal reinforcement learning.
In \emph{NeurIPS}, 2023.
\url{https://arxiv.org/abs/2303.11366}

\bibitem{chen2023typefly}
G.~Chen, X.~Yu, and L.~Zhong.
{TypeFly}: Flying drones with large language model.
\emph{arXiv preprint arXiv:2312.14950}, 2023.
\url{https://arxiv.org/abs/2312.14950}

\bibitem{star2025}
M.~Yuan, H.~Zhang, M.~Mohammadi, R.~Li, J.~Shan, and S.~L. Waslander.
{STaR}: Scalable task-conditioned retrieval for long-horizon multimodal robot memory.
\emph{arXiv preprint arXiv:2602.09255}, 2026.
\url{https://arxiv.org/abs/2602.09255}

\end{thebibliography}

\end{document}
